% ============================================================
%  LITERATURE REVIEW
% ============================================================
\section{Literature Review}

\subsection{Autoregressive Language Models and the GPT Lineage}

The modern era of large language models began with the Transformer architecture
\parencite{vaswani2017attention}, which replaced recurrent networks with self-attention and
positional encodings, enabling vastly better parallelisation during training.
GPT-2 \parencite{radford2019gpt2} demonstrated that a purely autoregressive, decoder-only
Transformer trained on a large web corpus could achieve strong zero-shot and few-shot performance
across a diverse set of NLP benchmarks without task-specific fine-tuning.
Subsequent scaling---GPT-3 \parencite{brown2020language} and beyond---showed that language
modelling loss and downstream task performance both improve smoothly as a power law in compute,
dataset size, and parameter count \parencite{hoffmann2022chinchilla}.
While this scaling paradigm has driven frontier capabilities, it has also substantially raised
the barrier to academic experimentation.

Recent efforts have therefore explored whether carefully designed \emph{small} models can close
the gap with brute-force scale.
Compact, reproducible frameworks such as NanoGPT \parencite{karpathy2022nanogpt} and its
successor NanoChat \parencite{nanochat} provide minimal, end-to-end implementations of GPT-2
style models that can be trained on a single GPU node for a cost of around \$100.
NanoChat in particular packages the model as a complete chatbot harness, including pretraining on
FineWeb-Edu \parencite{finewebedu,penedo2024fineweb}, a BPE tokeniser with UTF-8 byte fallback,
and evaluation against the DataComp-LM CORE suite \parencite{li2025datacomplmsearchgenerationtraining}.
These frameworks enable systematic, reproducible ablation studies that are infeasible at frontier
scale.

\subsection{Residual Connections and their Limitations}

Residual connections were introduced by \textcite{he2016deep} to address the vanishing gradient
problem in very deep networks, enabling the training of hundred-layer convolutional networks.
The update $h_l = h_{l-1} + f_{l-1}(h_{l-1})$ provides an identity shortcut that lets gradients
flow unimpeded to early layers.
Transformer architectures \parencite{vaswani2017attention} universally adopt a variant of this
pattern, typically combined with Pre-Layer Normalisation (PreNorm) for training stability.

However, \textcite{chen2026attnres} identify a structural limitation: standard residuals
accumulate all previous layer outputs with \emph{fixed, uniform weights}, causing the hidden
state magnitude to grow as $O(L)$ with depth and progressively diluting each individual layer's
contribution---a phenomenon they term \emph{PreNorm dilution}.
They propose \textbf{Attention Residuals (AttnRes)}, which replaces the fixed additive
accumulation with learned, input-dependent softmax attention over all preceding layer outputs.
This allows each layer to selectively retrieve relevant representations from any depth.
For large-scale training, they introduce Block AttnRes, which partitions layers into blocks and
attends over block-level aggregates, reducing memory from $O(Ld)$ to $O(Nd)$.
The Kimi Team demonstrates that AttnRes mitigates PreNorm dilution and improves training
dynamics, though their experiments are conducted at large scale.
Whether these gains transfer to shallow ($L=12$) models is an open question that we address
directly.

\subsection{Activation Functions}

The choice of activation function has a significant effect on the information flow and gradient
dynamics of deep networks.
The Rectified Linear Unit (ReLU) \parencite{nair2010rectified} became the default for many years
due to its simplicity and non-saturating gradient, though it suffers from the ``dying ReLU''
problem for negative inputs.
\textcite{hendrycks2016gelu} proposed the Gaussian Error Linear Unit (GeLU),
$\mathrm{GeLU}(x) = \frac{x}{2}\left[1 + \mathrm{erf}\!\left(\frac{x}{\sqrt{2}}\right)\right]$,
which provides a smooth, probabilistically-motivated gating of inputs.
GeLU has since become the dominant activation in large language models including GPT-2, BERT, and
their descendants.

NanoChat adopts $\mathrm{ReLU}^2$ (squaring the ReLU output), which has been shown to improve
sparsity and training efficiency in autoregressive models
\parencite{so2021primer}.
More recently, \textcite{vitvitskyi2026mininggeneralizableactivationfunctions} applied
evolutionary search to discover activation functions that generalise across tasks.
Their approach perturbs established functions with trigonometric components: \textit{GeLUSine}
adds a sinusoidal correction $0.1\sin(x)$ to GeLU, while \textit{GELU-Sinc-Perturbation} (GSP)
multiplicatively gates GeLU with a scaled sinc function.
These activations achieved improved performance on computer vision benchmarks with small models,
though their efficacy for language modelling had not been established.
Our experiments directly test this transfer.

\subsection{Multi-Token Prediction}

Standard next-token prediction trains the model to predict $x_{t+1}$ given the context
$x_{1:t}$.
While simple and effective, this one-step objective has been criticised for promoting
``lazy'' representations that exploit local statistics rather than longer-range structure, and is
vulnerable to \emph{error avalanche}: small mistakes in generation propagate and compound over
many steps.
\textcite{gloeckle2024better} proposed \textbf{Multi-Token Prediction (MTP)}, which trains
$n$ independent output heads to simultaneously predict the next $n$ tokens:
$L_n = -\sum_t \log P_\theta(x_{t+1:t+n} | x_{1:t})$.
The authors demonstrate that this objective encourages richer, more structured internal
representations, improving both perplexity and downstream task performance for large models
trained on substantial corpora.
MTP also opens additional inference-time options such as speculative decoding.
However, the overhead of $n$ output heads and the additional memory required to store intermediate
activations raise questions about its efficiency at academic scale---a gap we quantify in our
experiments.

\subsection{Memory-Augmented Language Models}

Augmenting Transformers with external or persistent memory is an active research direction aimed
at extending effective context beyond what fits in the attention window.
Early work such as Memory Networks \parencite{weston2015memory} and the Differentiable Neural
Computer \parencite{graves2016hybrid} demonstrated that explicit key-value stores can help models
reason over facts not present in a fixed context.
More recently, Retrieval-Augmented Generation (RAG) \parencite{lewis2020rag} has become a
standard technique for grounding LLMs in external knowledge at inference time, without modifying
model weights.

\textcite{cheng2026conditional} take a different approach with the \textbf{Engram layer}: a
conditional memory mechanism that augments individual transformer layers with a differentiable
key-value store addressed by n-gram hash IDs.
The module takes the current hidden state and n-gram context as input, performs an $O(1)$ lookup
to retrieve stored embeddings, and adds the result back into the residual stream before the
standard transformer block.
The original paper demonstrates consistent perplexity improvements on models of 27 billion
parameters trained on 262 billion tokens, framing the memory store as a new axis of sparsity
orthogonal to Mixture-of-Experts routing.
A key question is whether this mechanism, designed and validated at large scale, provides any
benefit for a shallow 12-layer model of approximately 85 million parameters trained for only 5000
steps---a significantly more data- and parameter-starved regime.

\subsection{Efficient Training: Mixed-Precision and FP8}

Beyond architectural modifications, training precision has emerged as an important efficiency
lever.
FP16 mixed-precision training \parencite{micikevicius2018mixed} became standard practice by
allowing most operations to run in half-precision while keeping a master copy of weights in
FP32 to preserve gradient fidelity.
More recently, the Hopper-generation H100 GPU introduced native FP8 tensor core support, enabling
8-bit floating-point matrix multiplications with theoretical throughput improvements.
\textcite{peng2023fp8} and subsequent work have shown that careful FP8 quantisation during
training can match FP16 performance while approximately doubling arithmetic throughput.
An emerging hypothesis---explored in our experiments---is that the quantisation noise introduced
by FP8 may act as an implicit regulariser, improving generalisation in data-limited regimes.

\subsection{Evaluation of Language Models at Small Scale}

Evaluating language models reliably at small scales is non-trivial.
Standard benchmarks designed for large models often show high variance when applied to models
below a certain capability threshold.
The CORE evaluation suite from DataComp-LM \parencite{li2025datacomplmsearchgenerationtraining}
addresses this by selecting 22 tasks specifically for their low-variance signal at small scales,
computing \emph{centered accuracies} to normalise for task difficulty.
We adopt CORE as our primary downstream metric, supplemented by validation BPB and tokens per
second as efficiency measures.
