% ============================================================
%  APPROACH / METHODOLOGY
% ============================================================
\section{Approach}

This project aims to improve NanoChat by incorporating and evaluating several recent modelling
and training innovations.
We use the existing NanoChat codebase \parencite{nanochat} as our starting point, which provides
a GPT-2-style transformer architecture, a distributed training loop, and an evaluation harness.
All four modifications described below are implemented by us on top of this codebase.
Each modification is tested in isolation as a controlled experiment against the unmodified
baseline, so that any performance change can be attributed to the specific technique.

\subsection{NanoChat Architecture}

NanoChat follows the GPT-2 decoder-only Transformer architecture with several modern refinements.
The model consists of a stack of 12 Transformer blocks, each containing a Multi-Query Attention
(MQA) sublayer followed by a Feed-Forward Network (FFN) with $\mathrm{ReLU}^2$ activation.
Each sublayer is preceded by RMSNorm and followed by a residual addition.
Input tokens are embedded via a learned embedding table and positional information is encoded
with rotary position embeddings (RoPE).
The model dimension is 768, with 12 attention heads of dimension 64, and the window pattern
\texttt{SSSL} (three sliding-window attention layers followed by one full-context layer) is
repeated across the depth.
The vocabulary is constructed using a custom Byte-Pair Encoder with UTF-8 byte fallback,
eliminating the need for an \texttt{<UNK>} token.
The final hidden state is projected to the vocabulary with a tied weight linear layer followed by
a softmax.

\subsection{Experiment 1: Activation Functions}

Activation functions are typically selected according to disciplinary convention rather than
systematic empirical comparison.
We aim to explicitly test several activation functions in the NanoChat FFN layers, including
functions recently proposed via evolutionary search \parencite{vitvitskyi2026mininggeneralizableactivationfunctions}.
In all experiments, the activation function is substituted uniformly across all FFN sublayers.
We evaluate the following six functions:

\begin{enumerate}
    \item \textbf{ReLU\(^2\)} (baseline): $\mathrm{ReLU}^2(x) = \max(0,x)^2$

    \item \textbf{GeLU} \parencite{hendrycks2016gelu}:
        \[\mathrm{GeLU}(x) = \frac{x}{2}\!\left[1 + \mathrm{erf}\!\left(\frac{x}{\sqrt{2}}\right)\right]\]

    \item \textbf{GeLUSine}: a sinusoidal perturbation of GeLU,
        \[\mathrm{GeLUSine}(x) = \mathrm{GeLU}(x) + 0.1\sin x\]

    \item \textbf{GELU-Sinc-Perturbation (GSP)}: a multiplicative sinc gate applied to GeLU,
        \[\mathrm{GSP}(x) = \mathrm{GeLU}(x)\!\left[1 + \alpha\cdot\mathrm{sinc}(x)\right]
          = \mathrm{GeLU}(x)\!\left[1 + \frac{\alpha\sin(\pi x)}{\pi x}\right], \quad \alpha = 0.5\]

    \item \textbf{Gaussian-Modulated Tangent Unit (GMTU)}: a custom activation combining a
        damped hyperbolic tangent with a linear leak,
        \[f(x) = \alpha\tanh(\beta x)\cdot e^{-\gamma x^2} + \lambda x,
          \quad \alpha{=}1.0,\; \beta{=}1.5,\; \gamma{=}0.2,\; \lambda{=}0.1\]

    \item \textbf{Turbulent}: a batch-wise activation that introduces a statistically-normalised
        sinusoidal perturbation on top of a log-scale base,
        \[f(\mathbf{x}) = \mathrm{sgn}(x_i)\ln(1+c|x_i|) + Ae^{-\frac{1}{2}z_i^2}\sin(\omega x_i),
          \quad z_i = \frac{x_i - \mu_\mathcal{B}}{\sigma_\mathcal{B}+\epsilon}\]
        where $\mu_\mathcal{B}$ and $\sigma_\mathcal{B}$ are the batch mean and standard
        deviation, and $c{=}0.5$, $A{=}0.2$, $\omega{=}2.0$, $\epsilon{=}10^{-6}$.
\end{enumerate}

\subsection{Experiment 2: Multi-Token Prediction Loss}

The standard next-token prediction objective is:
\begin{equation*}
    L_1 = -\sum_t \log P_\theta(x_{t+1} \mid x_{1:t})
\end{equation*}

This one-step training signal is prone to error avalanche during inference, since any single
prediction error compounds over subsequent steps.
We substitute this with the \textbf{Multi-Token Prediction (MTP)} loss
\parencite{gloeckle2024better}, which trains $n$ independent output heads to simultaneously
predict the next $n$ tokens from a shared backbone:
\begin{equation*}
    L_n = -\sum_t \log P_\theta(x_{t+1:t+n} \mid x_{1:t})
\end{equation*}
We experiment with $n=4$ prediction heads.
MTP encourages the model to develop more structured internal representations, as the shared hidden
state must support accurate prediction at multiple horizons simultaneously.

\subsection{Experiment 3: Engram Layer}

We integrate an \textbf{Engram layer} \parencite{cheng2026conditional} into the NanoChat
architecture while keeping total parameter count approximately fixed by reducing the depth of one
standard transformer block.
The Engram layer augments specific transformer blocks with a differentiable key-value memory
store addressed by n-gram hash identifiers.
At each forward pass, the module (i) computes n-gram hash IDs from the input token sequence, (ii)
performs an $O(1)$ lookup to retrieve cached embeddings from the memory store, and (iii) adds the
retrieved vectors back into the hidden state via a learned gating mechanism before the standard
transformer block.
Following the original paper's prescription for shallow models, we attach Engram modules to
layers 2 and 6 (out of 12), anticipating that these positions balance early contextual grounding
with mid-depth feature enrichment.

\subsection{Experiment 4: Attention Residuals}

Standard transformer residual connections apply a \emph{fixed, uniform} additive update at each
layer: $h_l = h_{l-1} + f_{l-1}(h_{l-1})$.
This causes uncontrolled growth in hidden-state magnitude with depth and progressively dilutes the
contribution of early layers \parencite{chen2026attnres}.
We replace this fixed accumulation with \textbf{Attention Residuals (AttnRes)}
\parencite{chen2026attnres}: each layer computes a learned softmax attention over all preceding
layer outputs, with a small learnable per-layer pseudo-query vector $w_l$.
Formally, let $\{h_0, h_1, \ldots, h_{l-1}\}$ denote the stored outputs of all preceding layers.
The AttnRes update is:
\begin{equation*}
    h_l = \sum_{i=0}^{l-1} \alpha_{i \to l}\, h_i + f_{l-1}(h_{l-1}),
    \quad \boldsymbol{\alpha} = \mathrm{softmax}(w_l^\top [h_0, \ldots, h_{l-1}])
\end{equation*}
For NanoChat's modest depth of 12 layers, Full AttnRes is computationally practical and we do not
employ the Block AttnRes variant.
The overhead is minimal, as the preceding layer outputs are already retained in memory during
backpropagation.
